Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2017-02-13 manuscript/supplemental_material.Rnw
use TSI to show that housekeeping genes are OK

>>>>

<<tissue_specificity_index_vs_housekeeping,echo=FALSE,message=FALSE>>=
tissue.specificity.index <- function(expression){
    expression <- expression[!is.na(expression)]
    expression <- expression[is.finite(expression)]
    if (length(expression)==0) {
        return(0)
    }
    if (max(expression)==0) {
        return(0)
    }
    return(sum(1-expression/max(expression))/
               (length(expression)-1))
}
tissue.specific.data <- fread("zcat ../data/rna_tissue.csv.zip")
tissue.specific.markers <-
    tissue.specific.data[,list(`Gene name`,
                               "Tissue Specificity Index"=
                                   tissue.specificity.index(as.numeric(Value))
                               ),
                         by=Gene][!duplicated(Gene),]
rm(tissue.specific.data)

setkey(tissue.specific.markers,"Gene name")
core.placenta.genes <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species),
                                     list("Median FPKM"=median(FPKM),
                                          "Gene Type"=ifelse(housekeeping,
                                                             "Housekeeping","Not Housekeeping"),
                                          housekeeping=housekeeping[1]
                                          ),
                                     by=human_name]
setkey(core.placenta.genes,"human_name")
core.placenta.genes.tsi <-
    tissue.specific.markers[core.placenta.genes[!duplicated(human_name)]]
rm(tissue.specific.markers,core.placenta.genes)
@

\begin{figure}
<<tissue_specificity_index_vs_median_fpkm,echo=FALSE,message=FALSE>>=
grid::pushViewport(grid::viewport(layout=grid::grid.layout(ncol=2,nrow=1,widths=c(1.5,1))))
print(ggplot(core.placenta.genes.tsi,
             aes(x=`Median FPKM`,
                 y=`Tissue Specificity Index`,
                 color=ifelse(housekeeping,"Housekeeping","Not\nHousekeeping")))
      + geom_point()
      + stat_ellipse()
      + ylab("Tissue Specificity Index")
      + xlab("Median FPKM")
      + scale_x_log10(labels=trans_format("log10", math_format(10^.x)))
      + scale_color_discrete("")
      + theme(legend.position="top"),
      vp=grid::viewport(layout.pos.row=1,layout.pos.col=1)
      )
print(ggplot(core.placenta.genes.tsi,
             aes(x=ifelse(housekeeping,"Housekeeping","Not Housekeeping"),
                 y=`Tissue Specificity Index`,
                 color=ifelse(housekeeping,"Housekeeping","Not\nHousekeeping")))
      + geom_boxplot(outlier.color=NA)
      + geom_jitter()
      + ylab("Tissue Specificity Index")
      + xlab("Gene Type")
      + scale_color_discrete("")
      + theme(legend.position="top"),
      vp=grid::viewport(layout.pos.row=1,layout.pos.col=2)
      )
grid::popViewport()
grid::grid.text(c("A","B"),
                y=unit(1,"npc")-unit(0.1,"cm"),
                x=unit(c(0,1.5/2.5),"npc")+unit(0.1,"cm"),
                gp=gpar(fontsize=24),
                just=c("left","top"))

@ 
\caption[Tissue specificity index with housekeeping]{Tissue
  specificity index and housekeeping genes. In general, the set of
  housekeeping genes identified by
  \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} has lower
  tissue specificity
  index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}
  than genes which are not tissue specific (calculated based on the
  Human Protein Atlas version 16.1 RNA-seq data
  \cite{Uhlén.ea2015:ProteomicsTissue-basedmapof}). Some of the genes
  with higher tissue specificity (\textit{NPC} and \textit{CISD1} have
  outlier expression in a single tissue, even though they are
  moderately expressed in every tissue.)}
\label{fig:tissue_specificity_index_with_housekeeping}
\end{figure}

<<tissue_specificity_index_of_placenta_core_genes,echo=FALSE,results="asis">>=
print(xtable(core.placenta.genes.tsi[order(-`Median FPKM`),
                                     list(`Gene name`,
                                          `Tissue Specificity Index`,
                                          `Median FPKM`,
                                          `Gene Type`
                                          )],
             digits=2,
             display=c("s","s","f","f","s"),
             caption=c("Tissue specificity index~\\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles} (calculated based on the Human Protein Atlas version 16.1 RNA-seq data \\cite{Uhlén.ea2015:ProteomicsTissue-basedmapof}) of Placenta Core Genes ordered by median expression with housekeeping status~\\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} indicated.",
                       "Tissue specificity of Placenta Core Genes"
                       ),
             label="tab:tissue_specificity_index_of_plaenta_core_genes"
             ),
      include.rownames=FALSE,
      booktabs=TRUE,
      math.style.exponents=TRUE,
      size="\\footnotesize",
      caption.placement="top",
      tabular.environment="longtable",
      floating=FALSE
      )
@
